Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-21 demystify.tex
Updates from Overleaf

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{indeed be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

@@

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall that CS PhD programs in the US do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internships at companies and research labs.

The first two years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for a job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\emph{MS requirement and PhD duration}:  as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}, CS PhD programs in the US do not require an MS degree.  In contrast, many other countries require having an MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

\emph{Project proposal}: in many countries, you have to choose a project and adviser \emph{during} the application process (e.g., you write a proposal to a potential adviser). You start your research right from the beginning.

In the US, you often start your PhD without an adviser or project and find them later. You will have two initial years to explore and find an adviser and research topic. And then you often explore projects with that adviser before settling on one (can take a couple of years).

\emph{Course work}: as mentioned above in the US you will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you also have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who has finished all course work and exams and only needs to write and defend their dissertation.}.

@@

In the US, PhD graduates often apply directly for faculty positions (postdoc for US graduates is no longer a popular option as it was before).

@@

Understandably students get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your prof. first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).

@@

\paragraph{Locally} Start looking for research opportunities at your institution.
If you did well or liked a class, you can check with the professor of that class for research opportunities.
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you still should contact them and ask.

There might be also special programs for undergraduate research from the university, e.g., GMU has the OSCAR program, UNL has UCARE, NSF has REUs (for US citizens and permanent residents).

@@

\begin{commentbox}[Vu:]
    I maintain a close relationship with my former professors and mentors. When there is a new event in my life (or theirs), I often send them an email or call them, e.g., when I get married, have a new baby, new job, etc. When I was having issues at UNL and considering leaving, I talked to them. I don't think this bothers them a bit; I think they are actually interested in these ``dramas'' in my life and would be offended if I didn't come to them.

@@

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (GMU). Before GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at GMU (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

>>>>

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall from \autoref{sec:msrequirement} that CS PhD programs in the US do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internships at companies and research labs.

@@

The first two years you typically take coursework (somewhat equivalent to an MS study), find an adviser, and learn how to do research.  The next 2--3 years you focus on your research, form a dissertation topic, and get results published. The last 1--2 years you continue to publish, write and defend your dissertation, and look for a job.
In many cases you might take a summer or two off to do an internship to get additional research opportunities.
The PhDComics figure above shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\emph{MS requirement and PhD duration}:  CS PhD programs in the US do not require an MS degree (as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}).  In contrast, many other countries require having an MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

\emph{Project proposal}: in many countries, you have to choose a project and adviser \emph{during} the application process (e.g., you write a proposal to a potential adviser). But this allows you to start your research right from the beginning.

In the US, you often start your PhD without an adviser or project and find them later. You will have two initial years to explore and find an adviser and research topic. After that you explore projects with that adviser before settling on one (can take a couple of years).

\emph{Course work}: as mentioned above in the US you will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who has finished all course work and exams and only needs to write and defend their dissertation.}.

@@

In the US, PhD graduates often apply directly for faculty positions. Postdoc for US graduates is no longer a popular option as it was before.

@@

Students understandbly get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your prof. first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide budget or computer equipment to incoming PhD students (e.g., a laptop).

@@

\paragraph{Locally} Start looking for research opportunities at your institution.
If you did well or liked a class, you can check with the professor of that class for research opportunities.
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you can still contact them and ask.

There might be also special programs for undergraduate research from the university, e.g., Mason has the OSCAR program, UNL has UCARE, and NSF has REUs (for US citizens and permanent residents).

@@

\begin{commentbox}[Vu:]
    I maintain a close relationship with my former professors and mentors. When there is a new event in my life (or theirs), I often send them an email or call them, e.g., when I get married, have a new baby, new job, etc. When I was having issues at UNL and considering leaving, I talked to my PhD advisers Deepak and Steph. I think this does not bother them a bit and that they are actually interested in these ``dramas'' in my life.

@@

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University. Before Mason, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at Mason and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at Mason (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.
